Density-dependent selection and the limits of relative fitness

### 2017-12-05 manuscript.tex
l

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others. Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope). First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp. 468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}. Yet, while this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$. In the examples from the previous paragraph, different types \textit{are} affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction \cite[PAGE NUMBER]{crow_1970}. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time \citep{messer_2016}, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}. Thus, it is important that we deepen our  understanding of the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

%The constant-$N$ relative fitness description of selection also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Adaptive dynamics provides a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density. However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of selection itself, which particularly critical for making sense of evolution at the genetic level.

%The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004} \citep{prout_1980}).

%in which time advances in discrete non-overlapping generations and fitness is interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}.

 %The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In particular, the $f_1$ and $f_2$ nullclines are not even defined.

Here we introduce a novel model of density-dependent population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. Our model is firmly grounded in population ecology, with fundamental parameters given by birth and death rates, and competitive ability. We show that this model can also be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.

Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}. This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.

%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless

Our model is essentially a density-dependent generalization of the classic ecological lottery model \cite{chesson_1981}. In the lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a weighted lottery \citep{sale_77}.

The classic lottery model assumes a saturated population with constant $N$, and a large number of propagules dispersing to each territory (the Wright-Fisher model makes a similar ``infinite propagule'' assumption to justify sampling with replacement). As such, the lottery model breaks down at low densities (few propagules dispersing to each territory). Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation''). We then...

@@

\subsection*{Basic assumptions}

@@

@@

In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference. It seems that Eq. \eqref{eq:master} or something similar (and hopefully more analytically tractable) is unavoidable for the analysis of time-course genetic data because, fundamentally, selective births and deaths affect both abundances and frequencies, not one or the other in isolation. Moreover, some aspects of allele frequency change are intrinsically density-dependent. In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model''). This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist. This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}. It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}). In cyclical environments, polymorphisms can be stabilized by the bounded density effect, which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}. We leave the details of how our model might be applied to inference problems, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.

>>>>

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope). First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp. 468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}. While this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$. In the examples from the previous paragraph, different types are affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather than starting from the standard models of population growth mentioned above, our point of departure  is the classic lottery model \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.

We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.

 Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation''). We then...

@@

\subsection*{Model assumptions and definitions}

@@

%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979}

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density. However, the focus of adatpive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here has been the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.
